\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-13 tex/paper.tex
spell check

\begin{abstract}
Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model to
continuous or discrete  data. \kcnote{emperical variational bayes}. We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

@@

In parallel, with the introduction of variational
autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\citep{2016arXiv161003483M}.  While these implicit generative networks
also do not admit a tractable density, they are differentiable.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness.
Unlike implicit generative models defined by neural networks, we do not assume
$\bfz$ to be a fixed-size vector with a simple density. Instead, the
dimensionality of $\bfz$ and the nature of its components (uniform, normal,
discrete, continuous, etc.) are inherited from the control flow of the
simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
the dimensionality of $\bfz$ may be much larger than the dimensionality of
$\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the marginal distribution parameterized by $\bfpsi$
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in \citep{2016arXiv161009033R}, though more complicated than a simple reparameterization of normally distributed noise $\bfz$ through a differentiable function $g(\bfz, \bfpsi)$. In our case, the variational program is a marginalized, non-differentiable  simulator.  Its density is  intractable; nevertheless, it can generate samples for $\bfx$ and is differentiable with respect to $\bfpsi$ and $\bfphi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

@@

Typically in variational inference one starts with a tractable prior
$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
approximate the posterior, one would criticize the model through the the
posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by
directly minimizing the  divergence of the corresponding posterior predictive
distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating
the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
Consequently, the proposed algorithm is a form of variational inference through
model criticism~\citep{box1980sampling}, which we denote AVI-MC. While the
likelihood is intractable, it is fixed, thus one can view the optimization of
$\bfpsi$ through the lens of emperical Bayes, though the prior family
$p(\bftheta | \bfpsi)$ is not explicitly specified.

@@

When $\rho$ is the Wasserstein distance, $\bftheta^*$ is refered to as the
minimum Wasserstein estimator. When the model is well specified, $\bftheta^*$
coincides with the true data-generating parameter; however, if the model is
misspecified, $\bftheta^*$ is typically different from the maximum likelihood
estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx |
\bftheta)$~\citep{bernton2017inference}. In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

As a first illustrative experiment, we evaluate inference for a discrete Poisson
distribution with unknown mean $\lambda$. We artificially consider
the distribution as a parameterized simulator, from which we can only
generate data.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Density $q(\bftheta|\bfpsi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization ($\gamma=10$). The proposal density correctly converges towards a distributon whose density concentrates around $\bftheta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

In this example, we consider observed data generated with the simplified generator of Eq.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored (mean field) Gaussian variational distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

@@

%Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
%which is a generalization of variational inference formulated as the following optimization problem
%\begin{eqnarray}
%\bflambda^* = \inf_\bfpsi \sup_\bflambda \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
%\end{eqnarray}
%In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
%AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
%\begin{eqnarray} \label{eq:opvi_avo}
%\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \\
%\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \;.
%\end{eqnarray}
%Rewriting Eq.~\ref{eq:opvi_avo} as Eq.~\ref{eq:opvi_avo2} is possible through importance sampling
%and reinforces the link to density ratio estimation as this corresponds to the operator
%\begin{equation}
%(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, .
%\end{equation}

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eq.~\ref{eq:opvi_avo} as Eq.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.

>>>>

\begin{abstract}
Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model to
continuous or discrete  data. \kcnote{empirical variational bayes}. We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

@@

In parallel, with the introduction of variational
auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\citep{2016arXiv161003483M}.  While these implicit generative networks
also do not admit a tractable density, they are differentiable.

@@

We consider a family of parametrized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness.
Unlike implicit generative models defined by neural networks, we do not assume
$\bfz$ to be a fixed-size vector with a simple density. Instead, the
dimension of $\bfz$ and the nature of its components (uniform, normal,
discrete, continuous, etc.) are inherited from the control flow of the
simulation code and may depend on $\bftheta$ in some intricate way. Moreover,
the dimension of $\bfz$ may be much larger than the dimension of
$\bfx$. Thus we do not attempt to learn a recognition model $q(\bfz | \bfx)$.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the marginal distribution parametrized by $\bfpsi$
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $p(\bfx|\bfpsi)$ as a \textit{variational program} as described in \citep{2016arXiv161009033R}, though more complicated than a simple reparameterization of normally distributed noise $\bfz$ through a differentiable function $g(\bfz, \bfpsi)$. In our case, the variational program is a marginalized, non-differentiable  simulator.  Its density is  intractable; nevertheless, it can generate samples for $\bfx$ and is differentiable with respect to $\bfpsi$ and $\bfphi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

@@

Typically in variational inference one starts with a tractable prior
$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
approximate the posterior, one would criticize the model through the the
posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by
directly minimizing the  divergence of the corresponding posterior predictive
distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating
the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
Consequently, the proposed algorithm is a form of variational inference through
model criticism~\citep{box1980sampling}, which we denote AVI-MC. While the
likelihood is intractable, it is fixed, thus one can view the optimization of
$\bfpsi$ through the lens of empirical Bayes, though the prior family
$p(\bftheta | \bfpsi)$ is not explicitly specified.

@@

When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
minimum Wasserstein estimator. When the model is well specified, $\bftheta^*$
coincides with the true data-generating parameter; however, if the model is
misspecified, $\bftheta^*$ is typically different from the maximum likelihood
estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx |
\bftheta)$~\citep{bernton2017inference}. In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

As a first illustrative experiment, we evaluate inference for a discrete Poisson
distribution with unknown mean $\lambda$. We artificially consider
the distribution as a parametrized simulator, from which we can only
generate data.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/multi.pdf}
    \caption{Multidimensional continuous data.
             ({\it Left}) Density $q(\bftheta|\bfpsi)$ at the beginning of the procedure, for a proposal distribution initialized with zero mean and unit variance.
             ({\it Right}) Density $q(\bftheta|\bfpsi)$ after adversarial variational optimization ($\gamma=10$). The proposal density correctly converges towards a distribution whose density concentrates around $\bftheta^* = (1, -1)$.
             }\label{fig:multi}
\end{figure}

@@

In this example, we consider observed data generated with the simplified generator of Eqn.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored (mean field) Gaussian variational distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

@@

%Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
%which is a generalization of variational inference formulated as the following optimization problem
%\begin{eqnarray}
%\bflambda^* = \inf_\bfpsi \sup_\bflambda \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
%\end{eqnarray}
%In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
%AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
%\begin{eqnarray} \label{eq:opvi_avo}
%\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \\
%\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \;.
%\end{eqnarray}
%Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling
%and reinforces the link to density ratio estimation as this corresponds to the operator
%\begin{equation}
%(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, .
%\end{equation}

Lastly, we make a connection to Operator Variational Inference (OPVI)~\citep{2016arXiv161009033R},
which is a generalization of variational inference formulated as the following optimization problem
\begin{eqnarray}
\bflambda^* = \inf_\bflambda \sup_\bfphi \mathbb{E}_{ \bfz \sim q( \bfz | \bflambda)} [ (O^{p,q} f_\bfphi) ] \, .
\end{eqnarray}
In traditional VI with the KL distance, this corresponds to $(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}$.
AVO can be cast into a similar form with expectations over $\bfx$ instead of $\bfz$ and
\begin{eqnarray} \label{eq:opvi_avo}
\bfpsi^* &=& \inf_\bfpsi \sup_\bfphi \mathbb{E}_{ \bfx \sim q( \bfx | \bfpsi)} [ (O^{p_r,p_\bfpsi} d_\bfphi) ] \\\
\label{eq:opvi_avo2} &=&  \inf_\bfpsi \sup_\bfphi\mathbb{E}_{\bfx\sim p_r(\bfx)}[d(\bfx)] - \mathbb{E}_{\bfx \sim q(\bfx|\bfpsi)}[d(\bfx)] \, .
\end{eqnarray}
Rewriting Eqn.~\ref{eq:opvi_avo} as Eqn.~\ref{eq:opvi_avo2} is possible through importance sampling,
corresponding to an implicit form of the operator
\begin{equation}
(O^{p,q} d) = \left( \frac{p_r(\bfx)}{q(\bfx|\bfpsi)}  -1 \right) d(\bfx) \, ,
\end{equation}
which reinforces the link to density ratio estimation and inference with implicit models.
